Supplemental Material for ``Assessing United States county-level exposure for research on tropical cyclones and human health"

### 2019-07-27 writing/ehp_methods.tex
Work on draft throughout

All counties in the eastern half of the \ac{US}
(Figure~\ref{fig:ivanexposure}), as of the~2010 \ac{US} Decennial Census, were
included in the study. The study covered all tracked storms in the \ac{HURDAT2}
\parencite{landsea2013} that came within~250 \si{\kilo\metre} of at least one
study county between~1988 and~2015 (Figure~S2\todo{Check fig ref}).
\ac{HURDAT2} is a post-storm assessment conducted by the \ac{US} \ac{NHC} and
incorporates data from a variety of sources, including satellite data and, when
available, aircraft reconnaissance data~\parencite{landsea2013, jarvinen1988}.
These data typically give measurements of the tropical cyclone center's
location at~6-\si{\hour} intervals at synoptic times (i.e., 6:00~am, 12:00~pm,
6:00~pm, and 12:00~am \ac{UTC}); some landfalling tropical cyclones have an
additional observation at the time of landfall~\parencite{landsea2013}.

@@

For each storm, the distance of closest approach was measured for every study
county.  Storm center track location data from \ac{HURDAT2} were interpolated
using natural cubic splines to~15-\si{\minute} intervals~\parencite{hurricaneexposure}.
The distance between each county's population mean center, as assessed by the
2010 Decennial Census~\parencite{countycenters}, and each~15-\si{\minute} interpolated
storm center location was measured using Great Circle (WGS84 ellipsoid)
distance~\parencite{bivand2013applied}. The distance-based exposure metric was
based on the minimum value of this storm center-to-county distance throughout
the period the storm was tracked. The time of closest approach for each
county-storm pairing was also recorded for use assessing rain-, flood-, and
tornado-based metrics. To allow matching with data based on local time, times
were converted from \ac{UTC} to local time~\parencite{countytimezones}.

@@

For the rain-based exposure metric, cumulative rainfall was estimated for the
period from two days before to one day after the date of a tropical cyclone's
closest approach to a county. Daily county-level rainfall was determined by
aggregating hourly~1/8\si{\degree} gridded reanalysis data from the \ac{NLDAS-2}
precipitation data files~\parencite{rui2013nldas}. The \ac{NLDAS-2} data integrate
satellite-based and land-based monitoring, applying a land-surface model to
create a reanalysis dataset that is spatially and temporally complete across
the continental \ac{US}~\parencite{rui2013nldas, alhamdan2014environmental}. To
aggregate to county-level, hourly data at each grid point were summed to create
a daily rainfall total, and these grid point rainfall totals were then averaged
across all grid points within a county's~1990 \ac{US} Census boundaries to
create a county-level daily total~\parencite{alhamdan2014environmental, cdcwonder}.
This county-level precipitation data are publicly available through the \ac{US}
\ac{CDC}'s \ac{WONDER} database~\parencite{cdcwonder, alhamdan2014environmental}.
The rainfall-based exposure metric was only available for tropical cyclones
through~2011, the period for which the county-aggregated reanalysis rainfall
data were available~\parencite{cdcwonder, alhamdan2014environmental}.

To assess the rainfall estimates from the reanalysis data used in analysis in
the the main text, we compared cumulative rainfall estimates based on this
reanalysis rainfall data to rainfall observations from ground-based weather
stations in nine sample study counties, selected as counties with data
available from multiple stations. In these nine counties, observed rainfall was
collected for all storms,~1988\,--\,2011, for the same period of two days
before to one day after each storm's closest approach, using all stations with
available daily data in the Global Historical Climatology Network throughout
1988\,--\,2011~\parencite{menne2012overview, rnoaa, countyweather}. Agreement
between measured storm-related rainfall between these two data sources was
assessed by Spearman's rank correlation.

@@

Ground-based observations of wind speed are problematic during tropical
cyclones, as instruments often fail at high wind speed, while reanalysis data,
often available at hourly or higher resolution, can be too smooth to capture
wind extremes associated with a tropical cyclone. Therefore, to create a
dataset of county-level sustained winds during historical tropical cyclones, we
modeled maximum sustained wind speeds at each county's population mean
center~\parencite{countycenters}. \ac{HURDAT2} Best Track data were interpolated
from~6-\si{\hour} reported values to~15-\si{\minute} increments using natural cubic
splines~\parencite{stormwindmodel}, and then sustained wind speed was modeled for
each county center at each~15-\si{\minute} increment using a double exponential wind
speed model~\parencite{willoughby2006parametric} to estimate maximum ground-level
sustained wind speed at each county center~\parencite{stormwindmodel}. Asymmetry in
wind speeds around the tropical cyclone center was incorporated into the wind
speed model~\parencite{phadke2003modeling, stormwindmodel}. The maximum value of
this modeled wind speed was determined for each county as the maximum windspeed
across the~15-\si{\minute} incremented modeled values throughout the tropical
cyclone.

To validate the metric of county-level wind speed used in the main text, we
compared this wind metric to estimated wind radii from \ac{HURDAT2}, available
for storms in our study from~2004~\parencite{landsea2013}. \ac{HURDAT2} gives
estimates of the maximum radii to winds of~64,~50, and~34 \si{\knot} in each four
quadrants of a storm (i.e., distance from the storm's center to the furthest
point in a quadrant with winds at or above that speed) at each~6-\si{\hour} storm
observation.

@@

To identify flood- and tornado-based exposures to tropical cyclones in \ac{US}
counties, we used event listings from the National Oceanic and Atmospheric
Administration (NOAA)'s Storm Event Database~\parencite{stormevents}. For each
tropical cyclone, we identified all events with event types related to flooding
(``Flood", ``Flash Flood", ``Coastal Flood") and tornadoes (``Tornado") and
that occurred in a county within~500 \si{\kilo\metre} of the tropical cyclone's track
and with a start date within a five-day window centered on the date of the
tropical cyclone's closest approach to the
county~\parencite{hurricaneexposuredata}. ``Flood", ``Flash Flood" and ``Tornado"
events in this database were reported by county \ac{FIPS} code. ``Coastal
Flood" events were reported by forecast zone; for these, the event was matched
to the appropriate county if possible~\parencite{noaastormevents}. While this
database has recorded storm data since~1950, its coverage changed substantially
in~1996 to cover a wider variety of storm events~\parencite{stormevents}. We
therefore only considered these metrics of hurricane exposure for tropical
cyclones in~1996 and later.

As validation for the flood data, we compared streamflow data from \ac{US}
Geological Survey gages~\parencite{usgsgages, countyfloods, dataRetrieval} in nine
counties with tropical cyclone flood classifications based on the NOAA Storm
Events Database (Figure~\ref{fig:floodcomparison}).

\subsection*{Assessing agreement in exposure classification between
metrics}

Two of the exposure metrics (flood- and tornado-based) were inherently binary,
since these metrics were based on whether an event was listed in the NOAA Storm
Events database or not.  For the other exposure metrics, each county was
classified as exposed to a tropical cyclone based on whether the exposure
metric exceeded a certain threshold (Table~\ref{tab:exposuremetrics}). We
assessed county-level tropical cyclone exposure in the eastern \ac{US} based on
each exposure metric for all land-falling or near-land Atlantic basin tropical
cyclones. Depending on available exposure data, this assessment included some
or all of the period from~1988 to~2015, a period with~136 tropical cyclones
that made \ac{US} landfall or passed within~250 \si{\kilo\metre} (155 mi) of at
least one \ac{US} county (Figure~S2\todo{Check fig ref}).

We measured agreement between exposure metrics in the classification of a
county as exposed or unexposed to a specific tropical cyclone. For this
measurement, we calculated the within-storm Jaccard
index~\parencite{jaccard1901distribution, jaccard1908nouvelles} between every
pair-wise combination of hazard metrics.  The Jaccard index~($J_s$) measures
similarity between two metrics ($X_{1,s}$ and~$X_{2,s}$) for tropical
cyclone~$s$ as the proportion of counties in which both of the metrics classify
the county as exposed out of all counties classified as exposed by at least one
of the metrics:

@@

\noindent This metric can range from~0, in the case of no overlap between the
counties classified as exposed based on the two metrics, to~1, in the case that
the two metrics classify exactly the same counties as exposed to the tropical
cyclone.

@@

The outcomes of disasters depend both on the geophysical forces of the disaster
as well as on the vulnerability of the society living in the geographical areas
affected by those forces.~\parencite{chakraborty2005population,
anderson2003community, cutter1996vulnerability} As a case study of how
differences in tropical cyclone exposure assessments across different hazard
metrics influence estimates of physical exposures of susceptible populations to
tropical cyclones, we investigated how the use of different metrics influenced
estimates of which eastern \ac{US} counties have the highest expected average
exposures of electricity-dependent Medicare beneficiaries to tropical cyclones.
This subpopulation was selected for this analysis since it may be particularly
susceptible to health impacts from tropical cyclone exposure, especially
through the pathway of storm-associated power outages and evacuations. We
collected data on the number of electricity-dependent Medicare beneficiaries in
each study county from the U.S. Department of Health \& Human Service's emPower
Map~2.0~\parencite{empower}. We then calculated the physical exposure of the
electricity-dependent Medicare population in each county, based on tropical
cyclone assessments using each hazard metric,
following~\parencite{peduzzi2009assessing}:

@@

\noindent where~$E_c$ is the average yearly physical exposure among
electricity-dependent Medicare beneficiaries in county~$c$ to tropical cyclone
exposures based on a given metric,~$F_c$ is the estimated yearly expected
frequency of tropical cyclone exposures in county~$c$ based on that metric,
and~$P_c$ is the size of the electricity-dependent Medicare population in
county~$c$, as of July~2017. When combined with estimates of vulnerability of a
population to a natural hazard, such measurements of physical exposure can be
used to calculate risk of human losses from the
hazard~\parencite{peduzzi2009assessing}.

>>>>

This study covered all counties in the eastern half of the \ac{US}
(Figure~\ref{fig:hurrtracks}), as of the~2010 \ac{US} Decennial Census. The
study covered all tracked land-falling or near-land Atlantic-basin tropical
cyclones between~1988 and~2015, considering all 136 storms in \ac{HURDAT2}
\parencite{landsea2013} that came within~250 \si{\kilo\metre} of at least one
eastern \ac{US} county in this period (Figure~\ref{fig:hurrtracks}).

\begin{comment}
These data typically give measurements of
the tropical cyclone center's location [and central pressure / maximum
windspeed?] at~6-\si{\hour} intervals at synoptic times (i.e., 6:00~am,
12:00~pm, 6:00~pm, and 12:00~am \ac{UTC}); some landfalling tropical cyclones
have an additional observation at the time of landfall~\parencite{landsea2013}.
\end{comment}

@@

We first measured how close each study storm came to each study county. We
started with tracking data from \ac{HURDAT2}, which records the storm's
position every~6 \si{\hour}, and interpolated to~15-\si{\minute} intervals
using natural cubic splines~\parencite{hurricaneexposure}. We then measured the
distance between the storm's position at each~15-\si{\minute} interval and each
county's population mean center, as assessed by the 2010 Decennial
Census~\parencite{countycenters}, using the Great Circle (WGS84 ellipsoid)
method~\parencite{bivand2013applied}. These measurements captured the distance
between the storm's central track and each study county throughout the period
when the storm was tracked. The closest distance between the storm and each
county was recorded as the distance of the storm's closest approach to the
county. We also recorded the time storm came closest to each county, for use in
assessing rain-, flood-, and tornado-based metrics where observed data on the
hazards needed to be matched by time to storm events.  To allow matching with
data based on local time, times were converted from \ac{UTC} to local
time~\parencite{countytimezones}.

@@

To estimate storm-associated rainfall, we used precipitation data from a
re-analysis dataset, as this data was available for every county and every day
through a continuous period of the study
(1988\,--\,2011)~\parencite{alhamdan2014environmental, cdcwonder}.  By
comparison, observations from ground-based monitoring networks had missing
values spatially (i.e., for some study counties), temporally (for some days),
or both.

We estimated daily rainfall for each study county by aggregating
hourly~1/8\si{\degree} gridded reanalysis data from the \ac{NLDAS-2}
precipitation data files~\parencite{rui2013nldas}. These data integrate
satellite-based and land-based monitoring, applying a land-surface model to
create a reanalysis dataset that is spatially and temporally complete across
the continental \ac{US}~\parencite{rui2013nldas, alhamdan2014environmental}. To
aggregate to county-level, the hourly data at each grid point were summed to
create a daily rainfall total, and these grid point rainfall totals were then
averaged across all grid points within a county's~1990 \ac{US} Census
boundaries~\parencite{alhamdan2014environmental, cdcwonder}. These daily
county-level estimates were then matched temporally with storm tracks, using
the date when each storm was closest to each county. The cumulative
storm-associated rainfall was then calculated as the sum of rainfall from two
days before the storm's closest approach to a county to one day after.

To validate these rainfall estimates, we investigated a subset of counties that
also had available ground-based precipitation observations. We selected nine
sample counties geographically spread across storm-prone regions of the eastern
\ac{US} and for which precipitation data were available from multiple stations
throughout the study period: Miami-Dade, FL; Harris County, TX; Mobile County,
AL; Orleans Parish, LA; Fulton County, GA; Charleston County, SC; Wake County,
NC; Baltimore County, MD; and Philadelphia County, PA. We collected
ground-based observations of precipitation data from all stations in the county
with available daily data in the Global Historical Climatology Network
throughout 1988\,--\,2011~\parencite{menne2012overview, rnoaa, countyweather}.
We summed daily station-specific measurments for two days before to one day
after each storm's closest approach and then averaged these cumulative
station-based precipitation totals for each county as a county-wide estimate of
cumulative storm-related precipitation. We measured the rank correlation
(Spearman's~$\rho$~\parencite{spearman1904proof}) between storm-specific
cumulative precipitation estimates for the two data sources within each sample
county.

@@

Ground-based observations of wind speed are problematic as a metric of exposure
to tropical cyclones, as instruments often fail at high wind speed, while
reanalysis data, often available at hourly or higher resolution, can lack a
fine enough temporal resolution to capture wind extremes associated with a
tropical cyclone. Therefore, to create a dataset of county-level sustained
winds during historical tropical cyclones, we modeled maximum sustained wind
speeds at each county's population mean center~\parencite{countycenters}, using
a wind speed model based on results from
Willoughby~\parencite{willoughby2006parametric}. For each storm, we used the
interpolated storm tracks generated for the distance metric and, for
each~15-\si{\minute} increment, we modeled maximum ground-level sustained wind
speed at each county center using a double exponential wind speed model with
inputs for the storm's forward speed, direction, and central wind
speed~\parencite{willoughby2006parametric, stormwindmodel}. This model
incorporated asymmetry in wind speeds around the tropical cyclone center that
results from the storm's forward movement~\parencite{phadke2003modeling,
stormwindmodel}. We then determined the maximum value of this estimate over the
storm tracking period for each study county to determine the maximum sustained
surface wind speed in each county during each storm.

To validate these modeled county-level wind speed estimates, we compared them
to county-level maximum sustained surface wind speeds based on the wind radii
estimates in \ac{HURDAT2}, which are based on \ldots. The wind radii values
have been included in \ac{HURDAT2} since 2004 [?] and estimate the distance
from the storm's center to the furthest point with winds at or above winds of a
certain speed in each four quadrants of the storm. They give estimates of these
maximum radii for three thresholds of maximum surface wind speeds: 64,~50,
and~34 \si{\knot}.  They therefore allow for the classification of counties
into four categories of maximum sustained wind speeds: $<$34~\si{\knot};
34\,--\,49.9~\si{\knot}; 50\,--\,63.9~\si{\knot}; $\ge$64~\si{\knot}. We
interpolated this data to~15-\si{\minute} increments and classified a county as
exposed to winds in a given wind speed category if its population mean center
was within~85\% of the maximum radius for that wind speed in its quadrant of
the storm [clarify why 85\%]. We compared these wind radii-based estimates with
the modeled wind speed estimates for the study storms since 2004 for which at
least one study county had a sustained wind of~$\ge$34 \si{\knot} (based on the
\ac{HURDAT2} wind radii). For each of these storms, we calculated the percent
of study counties that were classified in the same wind speed category.

@@

To identify flood- and tornado-based exposures to tropical cyclones in \ac{US}
counties, we used event listings from the National Oceanic and Atmospheric
Administration (NOAA)'s Storm Event Database~\parencite{stormevents}. For each
tropical cyclone, we identified all events with event types related to flooding
(``Flood", ``Flash Flood", ``Coastal Flood") and tornadoes (``Tornado") and
that occurred in a county within~500 \si{\kilo\metre} of the tropical cyclone's
track and with a start date within a five-day window centered on the date of
the tropical cyclone's closest approach to the
county~\parencite{hurricaneexposuredata}. ``Flood", ``Flash Flood" and
``Tornado" events in this database were reported by county \ac{FIPS} code.
``Coastal Flood" events were reported by forecast zone; for these, the event
was matched to the appropriate county if possible using regular expression
matching of listed county names~\parencite{noaastormevents}. While this
database has recorded storm data since~1950, its coverage changed substantially
in~1996 to cover a wider variety of storm events~\parencite{stormevents}. We
therefore only considered these metrics of hurricane exposure for tropical
cyclones in~1996 and later.

The tornado observations from this dataset form the traditional tornado
tracking dataset for the \ac{US} [ref for this?], and so we did not conduct
analyses to validate the tornado event data.  As validation for the
flood data, we investigated a sample of study counties, comparing the flood
status, based on NOAA Storm Events listings~\parencite{stormevents,
noaastormevents}, with streamflow measurements at \ac{US} Geological Survey county
streamflow gages during tropical storms~\parencite{usgsgages, countyfloods,
dataRetrieval}.  We considered nine study counties, selecting counties
geographically spread through storm-prone areas of the eastern \ac{US} and with
multiple streamflow gages reporting data during events (Baltimore County, MD;
Bergen County, NJ; Escambia County, FL; Fairfield County, CT; Fulton County,
GA; Harris County, TX; Mobile County, AL; Montgomery County, PL; and Wake
County, NC). For each county, we first identified all streamflow gages in the
county that had complete data for Jan.~1~1996\,--\,Dec.~31,~2015. For each
tropical cyclone that came within~500~\si{\kilo\metre} of one of these counties
between~1996 and~2015, we calculated the highest daily total streamflow (cubic
feet / second), summed across all identified streamgages in the county, for a
five-day window centered on the day of the storm's closest approach to the
county.  We also calculated the percent of streamflow gages in the county with
a daily streamflow that exceeded a threshold of flooding (the streamgage's
median value for annual peak flow~\parencite{countyfloods}) on any day during
the five-day window. We compared these measurements to each county's flood
status based on NOAA Storm Events listings for each tropical cyclone that came
within~500~\si{\kilo\metre} between~1996 and~2015, to explore if flood event
listings in the NOAA Storm Events database concurrent with tropical cyclones
were associated with higher streamflows at gages within the county.

\subsection*{Binary storm exposure classifications based on different metrics}

In our open-source dataset, we provide the closest distance of each storm to
each county, as well as county-level storm-related winds and rainfall, as
continuous metrics. However, epidemiologic studies of tropical cyclones often
compare ``exposed'' versus ``unexposed'' communities to assess storm-related
health risks [refs], and so we also used these continuous metrics to classify
counties as ``exposed'' or ``unexposed'' and explored patterns in storm
exposure based on these binary clasifications.

Two of the exposure metrics (flood- and tornado-based) were inherently binary,
since these metrics were based on whether an event was listed in the NOAA Storm
Events database.  For the other exposure metrics, each county was classified as
exposed to a tropical cyclone based on whether the exposure metric exceeded a
certain threshold (Table~\ref{tab:exposuremetrics}). For the rainfall metric, a
distance constraint was also necessary, to ensure that rains unrelated and far
from the storm track were not misattributed to a storm. Through exploratory
analysis, we set this distance metric at~500 \si{\kilo\metre} (i.e., for a
county to be classified as exposed based on rainfall, the cumulative rainfall
had to be over 75 \si{\milli\metre} and the storm must have passed within~500
\si{\kilo\metre} of the county; Table~\ref{tab:exposuremetrics}). We set this
distance constraint at a value that was typically large enough to capture
storm-related rain. However, data users should note that in rare examples of
exceptionally large storms (e.g., Hurricane Ike in 2008) or storms for which
storm tracking was stopped at extratropical transition (e.g., Tropical Storm
Lee in 2011), some storm-related rains may be missed because of this distance
constraint (Figure~S7).

We assessed patterns in county-level tropical cyclone exposure in the eastern
\ac{US} based on each exposure metric. Depending on available exposure data,
this assessment included some or all of the period from~1988 to~2015 (Table 1).
For each binary metric of tropical cyclone exposure (Table 1), we first summed
the total number of county-level exposures over available years for each
exposure metric and mapped patterns in these exposures. We next investigated
agreement between storm exposure classifications based on different exposure
metrics. We calculated the within-storm Jaccard
index~\parencite{jaccard1901distribution, jaccard1908nouvelles} between each
pair of exposure metrics. The Jaccard index~($J_s$) measures similarity between
two metrics ($X_{1,s}$ and~$X_{2,s}$) for tropical cyclone~$s$ as the
proportion of counties in which both of the metrics classify the county as
exposed out of all counties classified as exposed by at least one of the
metrics:

@@

\noindent This metric can range from~0, in the case of no overlap between the
counties classified as exposed based on the two metrics, to~1, in the case that
the two metrics classify exactly the same set of counties as exposed to the
tropical cyclone. We measured these values for all study storms that affected a
large number of eastern \ac{US} counties (250 or more of the study counties
classified as exposed by at least one of the exposure metrics) during the years
when all exposure data were available (1996\,--\,2011).

@@

Disasters' societal impacts depend both on the geophysical forces of the
disaster and on the vulnerability of those living in the affected geographical
areas~\parencite{chakraborty2005population, anderson2003community,
cutter1996vulnerability}. As a case study, we explored how differences in
tropical cyclone exposure assessments across different exposure metrics might
influence estimates of physical exposures of a susceptible population to
tropical cyclones. We calculated expected average physical exposures among
electricity-dependent Medicare beneficiaries in eastern \ac{US} counties to
tropical cyclones based on each exposure metric. This subpopulation was
selected since it may be particularly susceptible to health impacts from
tropical cyclone exposure, especially through the pathway of storm-associated
power outages and evacuations.

We collected data on the number of electricity-dependent Medicare beneficiaries
in each study county from the \ac{US} Department of Health \& Human Service's
emPower Map~2.0~\parencite{empower}. We then calculated the physical exposure
of the electricity-dependent Medicare population in each county, based on
tropical cyclone assessments using each hazard metric,
following~\parencite{peduzzi2009assessing}:

@@

\noindent where~$E_c$ is the average yearly physical exposure among
electricity-dependent Medicare beneficiaries in county~$c$ to tropical cyclone
exposures based on a given metric,~$F_c$ is the estimated yearly expected
frequency of tropical cyclone exposures in county~$c$ based on that metric,
and~$P_c$ is the electricity-dependent Medicare population in
county~$c$, as of July~2017.

\begin{comment}
When combined with estimates of vulnerability of a
population to a natural hazard, such measurements of physical exposure can be
used to calculate risk of human losses from the
hazard~\parencite{peduzzi2009assessing}.
\end{comment}
